Combining genomic prediction with citizen science for crop improvement and early selection in durum wheat

### 2019-06-28 manuscript/manuscript.Rmd
ms

Kauê de Sousa^1,2^, Jacob van Etten^2^, Carlo Fadda^3^, Yosef Gebrehawaryat^3^, Jean-Luc Jannink^4,5^, Basazen Fantahun Lakew^6^, Dejene Mengistu^7^, Svein Ø. Solberg^1^, Mario Enrico Pè^6^, Jesse Poland^8^, [… add others], Melfa and Workaye Farmer Community, Matteo Dell’Acqua^6[*]^  
^1^ Department of Agricultural Sciences, Inland Norway University of Applied Sciences, 2318 Hamar, Norway  
^2^ Bioversity International, Costa Rica Office, 30501 Turrialba, Costa Rica   
^3^ Bioversity International, Ethiopia Office, 1000 Addis Ababa, Ethiopia  
^4^ College of Agriculture and Life Sciences, Cornell University, 14853 Ithaca, USA  
^5^ Agricultural Research Service, United States Department of Agriculture, 148532901, Ithaca, USA  
^6^ Scuola Superiore Sant’Anna, Institute of Life Sciences, 56124 Pisa, Italy  
^7^ Bioversity International, Nepal Office, 3055 Kathmandu, Nepal  
^8^ Kansas State University, Kansas, USA  
^[*]^Correspondence should be addressed to MDA (email: <m.dellacqua@santannapisa.it>)

**Research questions:**

* Can farmers' evaluations better predict varietal performance associated with genes?
* Can we use crowdsourcing to improve GxE estimates?
* What is the ability of local varieties to outperform varieties from other zones?

# Introduction

* Crowdsourcing citizen science showed the ability to register the climate effects as they occur in farms, with potential to be used in plant breeding programs and risk assessments for farm adaptation[@VanEtten2019].

@@

## Model approaches

@@

# Results

# Discussion

* What is the relative costs of these approaches
* Is data missingness non-random? Can data missingness be predicted from farmer preferences? 
* Can we say something about which genomic regions play a role?

>>>>

Kauê de Sousa^1,2^, Jacob van Etten^2^, Svein Ø. Solberg^1^, Mario Enrico Pè^3^, Carlo Fadda^4^, Yosef Gebrehawaryat^4^, Jean-Luc Jannink^5,6^, Basazen Fantahun Lakew^3^, Dejene Mengistu^7^,  Jesse Poland^8^, [… add others], Matteo Dell’Acqua^3[*]^  
^1^ Department of Agricultural Sciences, Inland Norway University of Applied Sciences, 2318 Hamar, Norway  
^2^ Bioversity International, Costa Rica Office, 30501 Turrialba, Costa Rica   
^3^ Scuola Superiore Sant’Anna, Institute of Life Sciences, 56124 Pisa, Italy  
^4^ Bioversity International, Ethiopia Office, 1000 Addis Ababa, Ethiopia  
^5^ College of Agriculture and Life Sciences, Cornell University, 14853 Ithaca, USA  
^6^ Agricultural Research Service, United States Department of Agriculture, 148532901, Ithaca, USA  
^7^ Bioversity International, Nepal Office, 3055 Kathmandu, Nepal  
^8^ Kansas State University, Kansas, USA  
^[*]^Correspondence should be addressed to MDA (email: <m.dellacqua@santannapisa.it>)

# Research questions:

* Does farmers' evaluations associated with genes provide better predictions on varietal performance then station-only data?
* Does farmer's crowdsourced evaluations associated to SNP markers improve GxE estimates?
* Can we bring a new breeding paradigm where farmers actively participate in plant selection?

@@

## Genomic prediction

Methods from RBlup to genomic prediction with SNP markers. Farmers' evaluation from citizen science trials are connect to this approach by adding the probability of winning from Plackett-Luce as explanatory variables.

Other explanatory variables include the metric traits from station experiments.

## GxE interactions

Split varieties from citizen science trails by groups from their origin (cold tolerants and low regions). This cal also be used in the genomic prediction.

@@

# Topics for discussion (or keep in mind)

* GxE driven by under-representative samples or farm management.
* Is data missingness non-random? Can data missingness be predicted from farmer preferences? 
* Can we say something about which genomic regions play a role?
